---
title: "Terrain characteristics vs Land use in Fogo Island"
author: "Marcella Sarti Arellano"
date: "2024-09-10"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research question
1.a)
Does terrain characteristics (eg. exposition, slope) correlate with land cover, ie. are there major differences between single land covers?

# Libraries
```{r}
library(terra)
library(raster)
library(randomForest)
library(caret)
library(data.table)
library(sf)
library(dplyr)
library(vcd)
library(FactoMineR)
library(factoextra)
```

# 5. Relations with terrain characteristics

```{r}
# data
elevation<- rast("elevation_fogo.tif")

aspect <-rast("aspect_fogo.tif")

slope <- rast("slope_zona26.tif")

```

Extract values from the rasters

```{r}
# Elevation 
lcc2_resampled <- resample(lcc2, elevation, method = "near")

values_df <- as.data.frame(c(lcc2_resampled, elevation), na.rm = TRUE)
colnames(values_df) <- c("land_cover", "elevation")

values_df$land_cover <- as.factor(values_df$land_cover)
values_df$land_cover_num <- as.numeric(as.factor(values_df$land_cover))

# Aspect

lcc3_resampled <- resample(lcc2, aspect, method = "near")

values_df2 <- as.data.frame(c(lcc3_resampled, aspect), na.rm = TRUE)
colnames(values_df2) <- c("land_cover", "aspect")

values_df2$land_cover <- as.factor(values_df2$land_cover)
values_df2$land_cover_num <- as.numeric(as.factor(values_df2$land_cover))

# Slope

lcc4_resampled <- resample(lcc2, slope, method = "near")

values_df3 <- as.data.frame(c(lcc4_resampled, slope), na.rm = TRUE)
colnames(values_df3) <- c("land_cover", "slope")

values_df3$land_cover <- as.factor(values_df3$land_cover)
values_df3$land_cover_num <- as.numeric(as.factor(values_df3$land_cover))


```

## 5.0 Normality 

```{r}
# Elevation

hist(values_df$elevation, main="Elevation", xlab="elevation")

```
```{r}
# Aspect

hist(values_df2$aspect, main="Aspect", xlab="Aspect")

```

```{r}
# Elevation

hist(values_df3$slope, main="Slope", xlab="Slope")

```
Because of the lack of normality in elevation and slope, we use a non-parametric method: Kruskal-Wallis, which is adequate to test a categorical variable with a continuous.It evaluates the median of the slope/elevation within 2 or more groups of land cover. it's the non-parametric alternative for ANOVA.

For Aspect, because we have 2 categorical variables, we will use contingency tables, Cramer's and Fisher's test.

## 5.1. Elevation


```{r}
# Kruskal test
kruskal_elevation <- kruskal.test(elevation ~ land_cover, data = values_df)
kruskal_elevation
```
P-value < 0.05, elevation can differ with the type of land cover. But to see which of the land use may be the one that has the strongest relation we can visualize it on a box-plot.

```{r}
boxplot(elevation ~ land_cover, data = values_df, 
        xlab = "Land Cover", 
        ylab = "Elevation", 
        main = "Elevation & Land Cover",
        col = "lightblue", 
        las = 2)
```
Ocean and basaltic rock (old) may be the land cover types with the highest relation with elevation because of its small variability.

## 5.2. Aspect

```{r}
# names to aspect
values_df2 <- values_df2 %>%
  mutate(aspect2 = case_when(
    aspect == 2 ~ "N",
    aspect == 3 ~ "NE",
    aspect == 4 ~ "E",
    aspect == 5 ~ "SE",
    aspect == 6 ~ "S",
    aspect == 7 ~ "SW",
    aspect == 8 ~ "W",
    aspect == 9 ~ "NW",
    aspect == 10 ~ "N",
    TRUE ~ as.character(aspect)
  ))

# contingency table
table_cont<- table(values_df2$land_cover, values_df2$aspect2)
filtered_table <- table_cont[rowSums(table_cont) > 0, colSums(table_cont) > 0]


```

```{r}
# Correspondence analysis
ac <- CA(filtered_table, graph = FALSE)

summary(ac)

```
```{r}
# plot correspondence analysis
fviz_ca_biplot(ac, repel = TRUE)

```
p-value <0.05 = significant relation between the 2 variables

DIM 1:

On the right side of Dim.1: Categories such as "Basaltic rock (young)", "Forest", and columns like "NE" are found, suggesting an association of these areas with more natural and geologically young environments.

On the left side of Dim.1: Categories like "Land with little or no vegetation" are grouped, which is associated with arid or sparsely vegetated terrains, along with directions like "W" and "SW."

DIM 2:

Above Dim.2: "Human settlements" and "Basaltic rock (old)" are located.

Below Dim.2: "Agriculture" is found, implying that this dimension also distinguishes predominantly agricultural areas from other categories. "NW" and "S" are near "Agriculture," which could indicate some relationship or geographical proximity to cultivated areas.

Dim.1 seems to represent a gradient of vegetation, ranging from arid or sparsely vegetated areas to densely forested areas or young basaltic rocks.

Dim.2 captures a gradient of human intervention and geological age.


## 5.3. Slope
```{r}
# Kruskal test
kruskal_slope <- kruskal.test(slope ~ land_cover, data = values_df3)
kruskal_slope
```
P-value < 0.05, slope can differ with the type of land cover. But to see which of the land use may be the one that has the strongest relation we can visualize it on a box-plot.

```{r}
boxplot(slope ~ land_cover, data = values_df3, 
        xlab = "Land Cover", 
        ylab = "Slope", 
        main = "Slope & Land Cover",
        col = "lightblue", 
        las = 2)
```

Forest may be the land cover type with the highest relation with slope because of its small variability.
